---
title: "Example Applications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Applications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## Overview

This vignette demonstrates three applied analyses using `birdweatheR`: tracking 
migratory phenology at continental scale, characterizing activity timing in 
relation to the human footprint, and studying behavioral responses to the 
2024 total solar eclipse. Each example is self-contained and builds on the 
functions described in the [Getting Started vignette](getting-started.html).

```{r load_packages, eval = FALSE}
library(birdweatheR)
library(data.table)
library(ggplot2)

connect_birdweather()
```

---

## Migration Phenology

One of the most powerful applications of BirdWeather data is tracking the 
phenology of migratory species at continental scale. Because BirdWeather 
stations operate continuously, the dataset captures high-temporal-resolution 
information on the northward progression of spring migrants.

Here we use Wood Thrush (*Hylocichla mustelina*), a long-distance Neotropical 
migrant that winters in Central America and breeds across eastern North America, 
to illustrate how `get_detections()` can reveal migratory timing across 
latitudes.

The `woth` dataset bundled with this package contains 569,000 Wood Thrush 
detections from North American BirdWeather stations during spring migration 
(March–May 2025), filtered to detections with confidence ≥ 0.6.

```{r woth_load}
library(birdweatheR)
library(data.table)
library(ggplot2)

data(woth)
```

```{r woth_note, eval = FALSE}
# To reproduce this dataset from the API:
# connect_birdweather()
# woth_id <- find_species("Wood Thrush")$species_id
# woth <- get_detections(
#   from           = "2025-03-01T00:00:00.000Z",
#   to             = "2025-05-31T00:00:00.000Z",
#   species_ids    = woth_id,
#   confidence_gte = 0.6,
#   continents     = "North America"
# )
```

We calculate the median latitude of all detections for each day. As birds 
depart their Central American wintering grounds and move northward, the median 
latitude of detections shifts from roughly 15°N in early March to over 40°N 
by mid-May:

```{r woth_migration}
woth[, date := as.Date(timestamp)]

median_lat <- woth[, .(median_lat = median(station_lat, na.rm = TRUE)),
                   by = date]

ggplot(median_lat, aes(x = date, y = median_lat)) +
  geom_smooth(method = "gam", color = "#1C2223", fill = "#8D4C00") +
  annotate("text",
           x     = as.Date("2025-03-10"),
           y     = 20,
           label = "Wintering grounds\n(Central America)",
           size  = 3, color = "gray40") +
  annotate("text",
           x     = as.Date("2025-05-20"),
           y     = 45,
           label = "Breeding grounds\n(Eastern N. America)",
           size  = 3, color = "gray40") +
  labs(
    title    = "Wood Thrush Spring Migration",
    subtitle = "Northward shift in median detection latitude, March-May 2025",
    x        = NULL,
    y        = "Median Latitude (°N)",
  ) +
  theme_minimal()
```

The smooth northward shift from ~15°N to ~41°N over approximately ten weeks 
closely tracks published Wood Thrush migration chronology, demonstrating that 
BirdWeather acoustic detections can recover continent-scale movement patterns 
from passively collected data alone.

---

## Activity Timing Along a Human Footprint Gradient

Beyond tracking where birds are, BirdWeather's continuous monitoring enables 
examination of *when* birds are active — and how that timing varies across 
gradients of human disturbance. Here we use the bundled `woth` dataset to 
model the diel activity pattern of Wood Thrush across a gradient of the 
Human Footprint Index (HFI; Venter et al. 2016), a composite measure of 
anthropogenic pressure that integrates land use, human access, and 
infrastructure.

The response variable is binary: whether a given 30-minute interval at a 
given station contained at least one Wood Thrush detection. We model this 
probability as a function of time of day using harmonic (Fourier) terms 
within a binomial mixed model, with random effects for station. This 
approach — described in detail in Pease & Gilbert (2025) — captures the 
full shape of the diel activity curve rather than reducing it to a single 
peak hour.

### 1. Prepare station-level occasion histories

We start from the bundled `woth` detections, restricting to PUC stations 
that were active for at least 30 days across the March–May 2025 window. 
For each station, we construct a complete grid of 30-minute intervals and 
mark each interval as a detection (1) or non-detection (0):

```{r woth_hfi_prep, eval = FALSE}
library(lubridate)
library(dplyr)

# Restrict to PUC stations active for >= 30 days
woth[, date   := as.IDate(timestamp)]
woth[, n_days := uniqueN(date), by = station_id]
woth_puc <- woth[station_type == "puc" & n_days > 30]

# Parse timestamps and look up local timezone per station
d <- woth_puc[, .(timestamp, station_id, station_lat, station_lon)]
d[, zone      := lutz::tz_lookup_coords(station_lat, station_lon, method = "fast")]
d[, timestamp := lubridate::parse_date_time(timestamp, orders = "ymdHMSz")]
d[, date      := as.Date(timestamp)]
d <- d[!is.na(timestamp)]

station_tz <- d[, .(station_lat  = station_lat[1],
                    station_lon  = station_lon[1],
                    station_zone = zone[1]),
                by = station_id]

min_date <- min(d$date)
max_date <- max(d$date)

# Build complete 30-minute occasion grid per station
occasions <- vector("list", nrow(station_tz))
for (i in seq_len(nrow(station_tz))) {
  tz_i <- station_tz$station_zone[i]
  occasions[[i]] <- data.frame(
    station_id = station_tz$station_id[i],
    start = seq(
      from = lubridate::ymd_hms(paste(min_date, "00:00:00"), tz = tz_i),
      to   = lubridate::ymd_hms(paste(max_date, "00:00:00"), tz = tz_i),
      by   = "30 min"
    )
  ) |>
    dplyr::mutate(end = c(start[2:length(start)],
                          start[length(start)] + lubridate::minutes(30)))
}

occ <- data.table::rbindlist(occasions)
occ[, capt  := 0L]
occ[, start := lubridate::with_tz(start, "UTC")]
occ[, end   := lubridate::with_tz(end,   "UTC")]

# Mark intervals containing >= 1 detection
occ[d, capt := 1L,
    on = .(station_id,
           start <= timestamp,
           end   >  timestamp)]

# Interval midpoint and integer time-of-day bin (1-48)
occ[, mid      := start + as.numeric(difftime(end, start, units = "secs")) / 2]
occ[, mid_time := as.numeric(difftime(mid, lubridate::floor_date(mid, "day"),
                                       units = "mins"))]

# Summarise success/failure by station and time-of-day bin
final <- occ[, .(success = sum(capt),
                 failure = .N - sum(capt)),
             by = .(station_id, mid_time)]
final[, time := .GRP, by = mid_time]  # integer 1-48 circular predictor

# Join station coordinates
final <- station_tz[, .(station_id, station_lat, station_lon)][final, on = "station_id"]
```

### 2. Extract Human Footprint Index at station locations

The HFI raster (Venter et al. 2016) is downloaded once via `geodata` and 
values are extracted at each PUC station location using `terra`:

```{r woth_hfi_extract, eval = FALSE}
library(terra)
library(geodata)

# Download HFI raster (cached after first run)
hfi_rast <- geodata::footprint(year = 2009, path = tempdir())

# Extract HFI at station locations
stations_vect  <- terra::vect(as.data.frame(station_tz),
                               geom = c("station_lon", "station_lat"),
                               crs  = "EPSG:4326")
station_tz$hfi <- terra::extract(hfi_rast, stations_vect)[, 2]

# Join HFI onto modelling dataset
final <- final[station_tz[, .(station_id, hfi)], on = "station_id"]
```

### 3. Fit the circular mixed model

We fit a binomial GLMM with harmonic terms for time of day using 
`GLMMadaptive`. The fixed effects include two harmonics capturing daily 
and sub-daily cycles. To examine how activity timing varies with human footprint, HFI can be 
included as an interaction term in the fixed effects.Random effects allow each station's activity curve to deviate from the population mean. Model fitting takes several minutes 
depending on hardware:

```{r woth_hfi_model, eval = FALSE}
library(GLMMadaptive)

m1 <-  GLMMadaptive::mixed_model(
  fixed = cbind(success, failure) ~
    hfi * (
      cos(2 * pi * time / 48) + sin(2 * pi * time / 48) +
        cos(2 * pi * time / 24) + sin(2 * pi * time / 24)
    ),
  random = ~ cos(2 * pi * time / 48) + sin(2 * pi * time / 48) +
    cos(2 * pi * time / 24) + sin(2 * pi * time / 24) || station_id,
  data = final,
  family = binomial(),
  iter_EM = 0
)
```

### 4. Visualize the marginal activity curve

Marginal curves predicted at low and high HFI quantiles then reveal whether urbanization alters the shape or timing of the diel activity pattern. 

```{r woth_hfi_plot, eval = FALSE}
newdat <- expand.grid(
  time = seq(0, 48, length.out = 96), hfi = quantile(final$hfi, probs = c(0.1, 0.9))
)

marg_eff <- GLMMadaptive::effectPlotData(m2, newdat, marginal = TRUE) |>
  dplyr::mutate(across(pred:upp, plogis)) |>
  dplyr::mutate(time = time / 2)
marg_eff$hfi <- factor(marg_eff$hfi,
                       labels = c("Low human footprint", "High human footprint"))

ggplot(marg_eff, aes(
  x = time,
  y = pred,
  color = factor(hfi),
  fill = factor(hfi)
)) +
  
  geom_ribbon(aes(ymin = low, ymax = upp), alpha = 0.25, color = NA) +
  
  geom_line(linewidth = 1.5) +
  
  scale_x_continuous(limits = c(0, 24),
                     breaks = c(0, 4, 8, 12, 16, 20, 24)) +
  
  labs(x = "Time of Day (Hour)",
       y = "Probability of activity",
       color = "HFI",
       fill = "HFI") +
  
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black", linewidth = 0.2),
    axis.title = element_text(color = "black", size = 11),
    axis.text = element_text(color = "black", size = 10)
  )
```


---

## Astronomical Events

In addition to environmental sensors, PUCs record light levels at the sensor, 
including broadband light levels, near-infrared light, and the ability to isolate 
bands f1–f8. The combination of acoustic detections and on-board light sensors 
makes BirdWeather data uniquely suited to studying how birds respond to rapid 
changes in light levels — including those caused by solar eclipses.

### The April 8, 2024 Total Solar Eclipse

On April 8, 2024, a total solar eclipse crossed North America along a path 
running from Texas northeast through the Ohio Valley. At stations within the 
path of totality, daylight was completely extinguished for up to four minutes 
around 2:00 pm local time.

The `total_eclipse` dataset contains detections from PUC stations within an 
eclipse path bounding box during April 8, 2024. The `eclipse_light_data` dataset 
contains concurrent light sensor readings from those same stations.

```{r eclipse_load}
data(total_eclipse)
data(eclipse_light_data)
```

```{r eclipse_note, eval = FALSE}
# To reproduce from the API:
# connect_birdweather()
# total_eclipse <- get_detections(
#   from           = "2024-04-08T00:00:00.000Z",
#   to             = "2024-04-09T00:00:00.000Z",
#   ne             = list(lat = 42.639347, lon = -79.672852),
#   sw             = list(lat = 34.331340, lon = -94.570313),
#   confidence_gte = 0.6,
#   station_types  = "puc"
# )
#
# station_ids <- unique(total_eclipse$station_id)
# eclipse_light_data <- get_light_data(
#   station_id = station_ids,
#   from       = "2024-04-08T00:00:00.000Z",
#   to         = "2024-04-09T00:00:00.000Z"
# )
```

Rather than relying on an external shapefile, we use the PUC light sensor data 
itself to identify stations that experienced totality — defined here as stations 
where broadband light levels dropped below 10% of their morning baseline during 
the eclipse window:

```{r eclipse_totality}
eclipse_light_data[, datetime := as.POSIXct(timestamp,
                    format = "%Y-%m-%dT%H:%M:%S", tz = "America/Chicago")]
total_eclipse[, datetime := as.POSIXct(timestamp,
               format = "%Y-%m-%dT%H:%M:%S", tz = "America/Chicago")]

# Morning baseline (10:00-11:00 am)
station_baselines <- eclipse_light_data[
  datetime >= as.POSIXct("2024-04-08 10:00:00", tz = "America/Chicago") &
  datetime <= as.POSIXct("2024-04-08 11:00:00", tz = "America/Chicago"),
  .(baseline = mean(clear, na.rm = TRUE)), by = station_id]

# Minimum light during totality window (1:45-2:15 pm)
station_min <- eclipse_light_data[
  datetime >= as.POSIXct("2024-04-08 13:45:00", tz = "America/Chicago") &
  datetime <= as.POSIXct("2024-04-08 14:15:00", tz = "America/Chicago"),
  .(min_clear = min(clear, na.rm = TRUE)), by = station_id]

totality_ids <- merge(station_baselines, station_min, by = "station_id")[
  min_clear / baseline < 0.10, station_id]

message(length(totality_ids), " stations identified within path of totality")
```

Of the stations in the bounding box, 25 experienced a light drop of >90%, 
confirming they were likely within the path of totality.

We then compare broadband light levels with Common Grackle (*Quiscalus quiscula*) 
vocalization rates across these stations. Common Grackle is a highly vocal 
diurnal species with peak activity during midday — making it a sensitive 
indicator of unusual light changes:

```{r eclipse_5minbin}
# 5-minute bin helper
bin5 <- function(dt) {
  as.POSIXct(
    paste0(format(dt, "%Y-%m-%d %H:"),
           sprintf("%02d", (as.integer(format(dt, "%M")) %/% 5) * 5),
           ":00"),
    tz = "America/Chicago"
  )
}

# Filter to daytime window and totality stations
eclipse_day_light <- eclipse_light_data[
  station_id %in% totality_ids &
  datetime >= as.POSIXct("2024-04-08 10:00:00", tz = "America/Chicago") &
  datetime <= as.POSIXct("2024-04-08 17:00:00", tz = "America/Chicago")
]

eclipse_day_dets <- total_eclipse[
  station_id %in% totality_ids &
  datetime >= as.POSIXct("2024-04-08 10:00:00", tz = "America/Chicago") &
  datetime <= as.POSIXct("2024-04-08 17:00:00", tz = "America/Chicago") &
  common_name == "Common Grackle"
]

# Bin to 5 minutes
light_bins <- eclipse_day_light[,
  .(mean_clear = mean(clear, na.rm = TRUE)),
  by = .(bin = bin5(datetime))]

det_bins <- eclipse_day_dets[,
  .(det_count = .N),
  by = .(bin = bin5(datetime))]
```

```{r eclipse_plot}
library(cowplot)

p1 <- ggplot(det_bins, aes(x = bin, y = det_count)) +
  geom_line(color = "purple", alpha = 0.2) +
  geom_smooth(color = "purple", se = FALSE, span = 0.15) +
  geom_vline(
    xintercept = as.POSIXct("2024-04-08 13:59:00", tz = "America/Chicago"),
    color = "gray20", linetype = "dashed"
  ) +
  annotate("text",
           x     = as.POSIXct("2024-04-08 13:59:00", tz = "America/Chicago"),
           y     = max(det_bins$det_count, na.rm = TRUE) * 0.95,
           label = "Totality", color = "gray20", hjust = -0.1, size = 3) +
  labs(
    title    = "Common Grackle Vocalizations During the April 8, 2024 Solar Eclipse",
    subtitle = "5-minute detection counts, stations within path of totality",
    x        = NULL,
    y        = "Detections per 5 min"
  ) +
  theme_minimal()

p2 <- ggplot(light_bins, aes(x = bin, y = mean_clear)) +
  geom_line(color = "goldenrod") +
  geom_vline(
    xintercept = as.POSIXct("2024-04-08 13:59:00", tz = "America/Chicago"),
    color = "gray20", linetype = "dashed"
  ) +
  labs(
    subtitle = "Mean broadband clear light across stations in path of totality",
    x        = NULL,
    y        = "Clear Light (counts)",
    caption  = "Data source: BirdWeather via birdweatheR"
  ) +
  theme_minimal()

plot_grid(p1, p2, ncol = 1, align = "v")
```

The light sensor data clearly captures the eclipse — broadband light levels 
collapse to near zero at totality before recovering to normal afternoon levels. 
The concurrent reduction in Common Grackle vocalizations illustrates how 
BirdWeather's integrated acoustic and environmental sensing can be used to 
study behavioral responses to rapid environmental perturbations.

---

## Weather Events and Sensor Data

### The May 15–16, 2025 Tornado Outbreak

On May 15–16, 2025, a major tornado outbreak produced 61 tornadoes across the 
Midwest and Ohio Valley in the United States, including a destructive EF3 tornado 
that struck the Greater St. Louis metropolitan area at 2:39 pm CDT on May 16. 
The outbreak caused 26 deaths and approximately $5.9 billion in damages.

The `stl_storm_may2025` dataset contains detections from eight PUC stations in 
the St. Louis metro area during May 12–18, 2025. The `stl_env_may2025` dataset 
contains concurrent environmental sensor readings from those same stations.

```{r stl_load}
data(stl_storm_may2025)
data(stl_env_may2025)
```

```{r stl_note, eval = FALSE}
# To reproduce from the API:
# connect_birdweather()
# stl_storm_may2025 <- get_detections(
#   from          = "2025-05-12T00:00:00.000Z",
#   to            = "2025-05-18T00:00:00.000Z",
#   ne            = list(lat = 38.95, lon = -89.85),
#   sw            = list(lat = 38.35, lon = -90.75),
#   station_types = "puc"
# )
#
# station_ids <- unique(stl_storm_May2025$station_id)
# stl_env_may2025 <- get_environment_data(
#   station_id = station_ids,
#   from       = "2025-05-14T00:00:00.000Z",
#   to         = "2025-05-17T00:00:00.000Z"
# )
```

We focus on May 16 itself, plotting bird detections and mean temperature across 
all eight stations in 10-minute intervals. The sharp temperature drop associated 
with the tornado's passage is clearly visible, coinciding with a spike in 
acoustic detections around the time of EF3 touchdown:

```{r stl_may16}
# Parse detection timestamps
stl_storm_May2025[, datetime := as.POSIXct(timestamp,
                    format = "%Y-%m-%dT%H:%M:%S", tz = "America/Chicago")]

# 10-minute detection counts
ten_min_dets <- stl_storm_May2025[, .(det_count = .N),
                  by = .(interval = as.POSIXct(format(
                    datetime - as.numeric(datetime) %% 600,
                    "%Y-%m-%d %H:%M:%S"), tz = "America/Chicago"))]

stl_env_May2025[, datetime := as.POSIXct(timestamp,
                  format = "%Y-%m-%dT%H:%M:%S", tz = "America/Chicago")]

# 10-minute temperature averages
ten_min_env <- stl_env_May2025[, .(mean_temp = mean(temperature, na.rm = TRUE)),
                 by = .(interval = as.POSIXct(format(
                   datetime - as.numeric(datetime) %% 600,
                   "%Y-%m-%d %H:%M:%S"), tz = "America/Chicago"))]

# Merge and filter to May 16
ten_min_combined <- merge(ten_min_dets, ten_min_env, by = "interval", all = TRUE)
may16_start      <- as.POSIXct("2025-05-16 00:00:00", tz = "America/Chicago")
may16_end        <- as.POSIXct("2025-05-16 23:59:59", tz = "America/Chicago")
ten_min_may16    <- ten_min_combined[interval >= may16_start & interval <= may16_end]

# Intervals with no detections should be 0, not NA
ten_min_may16[is.na(det_count), det_count := 0]

library(cowplot)
p1 <- ggplot(ten_min_may16, aes(x = interval, y = det_count)) +
  geom_line(color = "forestgreen") +
  geom_vline(xintercept = as.POSIXct("2025-05-16 14:39:00", tz = "America/Chicago"),
             color = "red", linetype = "dashed") +
  labs(title = "Bird Detections — St. Louis Metro (May 16, 2025)",
       x = NULL, y = "Detections per 10 min") +
  theme_minimal()

temp_max <- max(ten_min_may16$mean_temp, na.rm = TRUE)
if (!is.finite(temp_max)) temp_max <- 25

p2 <- ggplot(ten_min_may16, aes(x = interval, y = mean_temp)) +
  geom_line(color = "darkorange") +
  geom_vline(xintercept = as.POSIXct("2025-05-16 14:39:00", tz = "America/Chicago"),
             color = "red", linetype = "dashed") +
  annotate("text", x = as.POSIXct("2025-05-16 14:39:00", tz = "America/Chicago"),
           y     = 30,
           label = "EF3 touchdown\n2:39pm CDT",
           color = "red", hjust = -0.1, size = 3) +
  labs(x = NULL, y = "Temperature (°C)",
       caption = "Data source: BirdWeather via birdweatheR") +
  theme_minimal()

plot_grid(p1, p2, ncol = 1, align = "v")
```

The `get_environment_data()` function retrieves sensor readings for any PUC 
station and time window, enabling researchers to pair environmental context with 
acoustic detections at fine temporal resolution.

---
## Further Reading

For an example of large-scale ecological synthesis using BirdWeather detections, 
see [Pease & Gilbert (2025)](https://www.science.org/doi/10.1126/science.adv9472), 
who used over 60 million BirdWeather detections across 583 diurnal species to 
demonstrate that light pollution prolongs avian vocal activity by up to 50 
minutes in the brightest landscapes (*Science*, 387, 848–853).

Additionally, a full analysis of the 2023 Annular and 2024 Total Eclipses using 
data from the [Eclipse Soundscapes](https://eclipsesoundscapes.org/) project and 
BirdWeather can be found in [Gilbert et al., 2026](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.73090).

For questions, bug reports, or feature requests, please visit the package 
repository at [github.com/BrentPease1/birdweatheR](https://github.com/BrentPease1/birdweatheR).
