---
title: "Getting Started with birdweatheR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with birdweatheR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## Introduction

[BirdWeather](https://birdweather.com) is a global network of acoustic monitoring 
stations maintained and operated by volunteers that continuously record and automatically identify bird vocalizations using the BirdNET neural network. As of February 2026, the network comprises over 18,000 stations across six continents, collectively generating tens of millions of detections per month. Just recently, the database hit 2 billion detections! Unlike citizen science platforms that rely on human observers, BirdWeather stations operate continuously and autonomously — making the dataset uniquely suited to studying fine-grained temporal patterns in bird behavior, including responses to weather events, light cycles, and astronomical phenomena. 

Currently, BirdWeather offers three options for accessing the database: 1) A user-friendly GUI data explorer, 2) GraphQL API, and 3) REST API. While the latter two are particularly powerful, not all users are comfortable with API calls. `birdweatheR` provides an R interface to the BirdWeather API, enabling researchers to download detection data, explore species activity patterns, andintegrate on-board sensor readings from BirdWeather PUC units — which record temperature, barometric pressure, humidity, air quality, and spectral light levels alongside acoustic detections.

### Installation

```{r install, eval = FALSE}
# Install from GitHub
# install.packages("devtools")
devtools::install_github("BrentPease1/birdweatheR")
```

### Connecting to the API

All functions require an active API connection. No API KEY required. Establish a connection at the start of 
each session:

```{r connect, eval = FALSE}
library(birdweatheR)
connect_birdweather()
```

The connection is stored in a package-level environment and used automatically by all downstream functions. You do not need to pass it explicitly after the initial call.

---

## Exploring the Platform

Before pulling raw detections, a few summary functions give a quick overview of the BirdWeather database.

### Platform-wide summary

`get_counts()` returns a single-row summary of detections, species, and stations for a given time period:

```{r get_counts, eval = FALSE}
get_counts(
  from = "2025-05-01T00:00:00.000Z",
  to   = "2025-05-02T00:00:00.000Z"
)
```

```{r get_counts_output, echo = FALSE}
data.frame(
  detections = 11190893,
  species    = 1664,
  stations   = 3890
)
```

Over 11 million detections from 1,664 species across 3,890 stations in a single 24-hour period illustrates the scale of the BirdWeather network.

### Top species

`get_top_species()` returns the most frequently detected species for a given period, with a breakdown of detection confidence:

```{r get_top_species, eval = FALSE}
get_top_species(
  limit = 10,
  from  = "2025-05-01T00:00:00.000Z",
  to    = "2025-05-02T00:00:00.000Z"
)
```

```{r get_top_species_output, echo = FALSE}
data.frame(
  species_id = c(1, 35, 134, 51),
  common_name    = c(
    "House Sparrow",
    "Eurasian Blackbird",
    "House Finch",
    "Common Chiffchaff"
  ),
  scientific_name   = c(
    "Passer domesticus",
    "Turdus merula",
    "Haemorhous mexicanus",
    "Phylloscopus collybita"
  ),
  count = c(829043, 713296, 519312, 436480),
  almost_certain = c(828755, 713294, 519242, 436477),
  very_likely = c(88, 2, 70, 3),
  unlikely = c(0, 0, 0, 0),
  uncertain = c(0, 0, 0, 0)
)
```

### Discovering stations

`get_stations()` retrieves the full station network with coordinates, country, and station type. This is useful for identifying stations in a region of interest before pulling detections:

```{r get_stations, eval = FALSE}
# Pull 10 stations
stations <- get_stations(limit = 10)

# Stations within a bounding box and time period
midwest_stations <- get_stations(
  from  = "2025-05-01T00:00:00.000Z",
  to    = "2025-05-02T00:00:00.000Z"
  ne = list(lat = 49.0, lon = -80.0),
  sw = list(lat = 36.0, lon = -97.0)
)
```

`midwest_stations` returns 569 observations of 10 variables, including station name, type (e.g., "PUC", see below), longitude and latitude, and a station_id unique to the database that could be helpful for other queries. 

---

## Species Detection Patterns

### Finding species

`find_species()` searches the BirdWeather species database by common or 
scientific name. This function allows for complete and partial matches, even including wildcards (e.g., 'whip*')

```{r find_species, eval = FALSE}
# Search by common name
find_species("Wood Thrush")

# Partial match
find_species("thrush")

# wildcard match
find_species("whip*")
```

### Time-of-day activity

`get_tod_counts()` returns detection counts binned by 30-minute intervals across a time period, revealing the diel activity pattern of a species. Here we examine the American Robin (*Turdus migratorius*) — a species well known for its early dawn and late evening chorus:

```{r tod_counts, eval = FALSE}
robin_id <- find_species("American Robin")$species_id

robin_tod <- get_tod_counts(
  species_id = robin_id,
  from       = "2025-05-01T00:00:00.000Z",
  to         = "2025-05-31T00:00:00.000Z"
)
```

```{r tod_plot, eval = FALSE}
library(ggplot2)

ggplot(robin_tod, aes(x = hour, y = count)) +
  geom_col(fill = "tomato") +
  scale_x_continuous(breaks = seq(0, 24, by = 3),
                     labels = paste0(seq(0, 24, by = 3), ":00")) +
  labs(
    title    = "American Robin — Time of Day Activity",
    subtitle = "May 2025, all BirdWeather stations",
    x        = "Hour of Day",
    y        = "Total Detections"
  ) +
  theme_minimal()
```

Contrasting that with a nocturnal species such as the Barred Owl (*Strix varia*) shows the power of continuous detections, creating a contrast with traditional sampling techniques and other avian databases:
```{r baow_tod_counts, eval = FALSE}
baow_id <- find_species("Strix varia")$species_id

baow_tod <- get_tod_counts(
  species_id = baow_id,
  from       = "2025-05-01T00:00:00.000Z",
  to         = "2025-05-31T00:00:00.000Z"
)
```

```{r baow_plot, eval = FALSE}
library(ggplot2)

ggplot(baow_tod, aes(x = hour, y = count)) +
  geom_col(fill = "blue4") +
  scale_x_continuous(breaks = seq(0, 24, by = 3),
                     labels = paste0(seq(0, 24, by = 3), ":00")) +
  labs(
    title    = "Barred Owl — Time of Day Activity",
    subtitle = "May 2025, all BirdWeather stations",
    x        = "Hour of Day",
    y        = "Total Detections"
  ) +
  theme_minimal()
```

One shortcoming of the TOD function, however, is when a species is poorly detected. In this case, a manual approach would be required (e.g., using `get_detections`)

### Daily detection trends

`get_daily_detection_counts()` returns detection totals aggregated by day, 
optionally broken down by species. This is useful for visualizing seasonal 
trends without downloading raw detections:

```{r daily_counts, eval = FALSE}
daily <- get_daily_detection_counts( #ART 20 seconds
  from       = "2025-05-01T00:00:00.000Z",
  to         = "2025-05-31T00:00:00.000Z",
  by_species = FALSE
)

ggplot(daily, aes(x = as.Date(date), y = daily_total)) +
  geom_smooth(color = "steelblue", se = F) +
  labs(
    title = "Daily BirdWeather Detections",
    x     = NULL,
    y     = "Detections per Day"
  ) +
  theme_minimal()
```

This function can be a powerful summary function when combined with `station_ids`, `species_ids`, or breaking down `by_species`, which are all built-in arguments to the function. 

### Raw Detections

For research, the most useful function will likely be `get_detections`. This function allows for raw data downloads from the BirdWeather database, where then the scientific questions are seemingly endless.

#### get_detections

`get_detections` is a function that retrieves bird detections from the BirdWeather API with optional filters. Importantly, since we are dealing with millions of rows, the function handles pagination automatically up to the user-specified limit. If no limit is specified, the function will return the total number of detections matching the query. Returns a fully flat data.table with all nested fields (coords, species, station) expanded into individual columns.

Current functionality allows for filtering on:
* Dates (beginning and end)
* Stations
* Types of Stations
* Species
* Continents
* Countries
* BirdNET Confidence Scores
* Bounding Box
* Download Limits (e.g., 1000 rows)

```{r get_detections, eval = FALSE}
# Get detections for a date range
detections <- get_detections(
  from  = "2025-01-01T00:00:00.000Z",
  to    = "2025-01-02T00:00:00.000Z",
  limit = 1000
)

# Filter by species name - if multiple matches are found, all are shown
# and the user is prompted to rerun with a more specific name
detections <- get_detections(
  from          = "2025-05-01T00:00:00.000Z",
  to            = "2025-05-02T00:00:00.000Z",
  species_names = "Eastern Whip-poor-will",
  limit         = 1000
)

# For faster pulls, look up species ID first and pass it directly
ewpw_id <- find_species("Eastern Whip-poor-will")$species_id

detections <- get_detections(
  from        = "2025-05-01T00:00:00.000Z",
  to          = "2025-05-02T00:00:00.000Z",
  species_ids = ewpw_id,
  limit       = 1000
)

# Filter by continent with a confidence threshold
detections <- get_detections(
  from           = "2025-01-01T00:00:00.000Z",
  to             = "2025-01-02T00:00:00.000Z",
  continents     = "North America",
  confidence_gte = 0.9,
  limit          = 1000
)

```

##### returned columns
`get_detections` will return the following columns:

| Column | Description |
|---|---|
| `id` | Detection ID |
| `timestamp` | Detection timestamp in station timezone |
| `confidence` | BirdNET confidence |
| `score` | Calculated identification likelihood score |
| `species_id` | Species ID |
| `common_name` | Species common name |
| `scientific_name` | Species scientific name |
| `station_id` | Station ID |
| `station_name` | Station name |
| `station_type` | Station type |
| `station_country` | Country |
| `station_continent` | Continent |
| `station_state` | State / province |
| `station_lat`, `station_lon` | Station coordinates |
| `location_privacy` | fuzzy lat/lons |

---

# Applications of BirdWeather

## Migration Phenology

One of the most powerful applications of BirdWeather data is tracking the 
phenology of migratory species at continental scale. Because BirdWeather 
stations operate continuously, the dataset captures the high-temporal resolution information, such as northward progression of 
spring migrants.

Here we use Wood Thrush (*Hylocichla mustelina*), a long-distance Neotropical 
migrant that winters in Central America and breeds across eastern North America, 
to illustrate how `get_detections()` can reveal migratory timing across 
latitudes.

The `woth` dataset bundled with this package contains 569,000 Wood Thrush 
detections from North American BirdWeather stations during spring migration 
(March–May 2025), filtered to detections with confidence ≥ 0.6.

```{r woth_load}
library(birdweatheR)
library(data.table)
library(ggplot2)

data(woth)
```

```{r woth_note, eval = FALSE}
# To reproduce this dataset from the API:
# connect_birdweather()
# woth_id <- find_species("Wood Thrush")$species_id
# woth <- get_detections(
#   from           = "2025-03-01T00:00:00.000Z",
#   to             = "2025-05-31T00:00:00.000Z",
#   species_ids    = woth_id,
#   confidence_gte = 0.6,
#   continents     = "North America"
# )
```

We calculate the median latitude of all detections for each day. As birds 
depart their Central American wintering grounds and move northward, the median 
latitude of detections shifts from roughly 15°N in early March to over 40°N 
by mid-May:

```{r woth_migration}
woth[, date := as.Date(timestamp)]

median_lat <- woth[, .(median_lat = median(station_lat, na.rm = TRUE)),
                   by = date]

ggplot(median_lat, aes(x = date, y = median_lat)) +
  geom_smooth(color = "#1C2223", fill = "#8D4C00") +
  annotate("text",
           x     = as.Date("2025-03-10"),
           y     = 20,
           label = "Wintering grounds\n(Central America)",
           size  = 3, color = "gray40") +
  annotate("text",
           x     = as.Date("2025-05-20"),
           y     = 45,
           label = "Breeding grounds\n(Eastern N. America)",
           size  = 3, color = "gray40") +
  labs(
    title    = "Wood Thrush Spring Migration",
    subtitle = "Northward shift in median detection latitude, March-May 2025",
    x        = NULL,
    y        = "Median Latitude (°N)",
    caption  = "Data source: BirdWeather via birdweatheR"
  ) +
  theme_minimal()
```

The smooth northward shift from ~15°N to ~41°N over approximately ten weeks 
closely tracks published Wood Thrush migration chronology, demonstrating that 
BirdWeather acoustic detections can recover continent-scale movement patterns 
from passively collected data alone.

---

## Weather Events and Sensor Data

[BirdWeather PUC (Portable Universe Codec) units](https://www.birdweather.com/shop-birdweather-puc), BirdWeather's AI-powered bioacoustics platform, in addition to being equipped with dual microphones, WiFi, and GPS, are also packaged with on-board environmental sensors that record 
temperature, humidity, barometric pressure, air quality index, and sound 
pressure level at approximately 42-second intervals. This onboard sensor data 
enables researchers to directly link bird vocal activity to local environmental 
conditions — a capability not available in any other large-scale acoustic 
monitoring dataset.

### The May 15–16, 2025 Tornado Outbreak
On May 15–16, 2025, a major tornado outbreak produced 61 tornadoes across the 
Midwest and Ohio Valley in the United States, including a destructive EF3 tornado that struck the Greater St. Louis metropolitan area at 2:39 pm CDT on May 16. The outbreak 
caused 26 deaths and approximately $5.9 billion in damages.

The `stl_storm_May2025` dataset contains detections from eight PUC stations in 
the St. Louis metro area during May 12–18, 2025. The `stl_env_May2025` dataset 
contains concurrent environmental sensor readings from those same stations.
```{r stl_load}
data(stl_storm_May2025)
data(stl_env_May2025)
```
```{r stl_note, eval = FALSE}
# To reproduce from the API:
# connect_birdweather()
# stl_storm_May2025 <- get_detections(
#   from          = "2025-05-12T00:00:00.000Z",
#   to            = "2025-05-18T00:00:00.000Z",
#   ne            = list(lat = 38.95, lon = -89.85),
#   sw            = list(lat = 38.35, lon = -90.75),
#   station_types = "puc"
# )
#
# station_ids <- unique(stl_storm_May2025$station_id)
# env_list <- lapply(station_ids, function(sid) {
#   get_environment_data(
#     station_id = sid,
#     from       = "2025-05-14T00:00:00.000Z",
#     to         = "2025-05-17T00:00:00.000Z"
#   )
# })
# stl_env_May2025 <- data.table::rbindlist(env_list, fill = TRUE)
```

We focus on May 16 itself, plotting bird detections and mean temperature across 
all eight stations in 10-minute intervals. The sharp temperature drop associated 
with the tornado's passage is clearly visible, coinciding with a spike in 
acoustic detections around the time of EF3 touchdown:
```{r stl_may16}
# Parse detection timestamps
stl_storm_May2025[, datetime := as.POSIXct(timestamp,
                    format = "%Y-%m-%dT%H:%M:%S", tz = "America/Chicago")]

# 10-minute detection counts
ten_min_dets <- stl_storm_May2025[, .(det_count = .N),
                  by = .(interval = as.POSIXct(format(
                    datetime - as.numeric(datetime) %% 600,
                    "%Y-%m-%d %H:%M:%S"), tz = "America/Chicago"))]

stl_env_May2025[, datetime := as.POSIXct(timestamp,
                  format = "%Y-%m-%dT%H:%M:%S", tz = "America/Chicago")]

# 10-minute temperature averages
ten_min_env <- stl_env_May2025[, .(mean_temp = mean(temperature, na.rm = TRUE)),
                 by = .(interval = as.POSIXct(format(
                   datetime - as.numeric(datetime) %% 600,
                   "%Y-%m-%d %H:%M:%S"), tz = "America/Chicago"))]

# Merge and filter to May 16
ten_min_combined <- merge(ten_min_dets, ten_min_env, by = "interval", all = TRUE)
may16_start      <- as.POSIXct("2025-05-16 00:00:00", tz = "America/Chicago")
may16_end        <- as.POSIXct("2025-05-16 23:59:59", tz = "America/Chicago")
ten_min_may16    <- ten_min_combined[interval >= may16_start & interval <= may16_end]

# Intervals with no detections should be 0, not NA
ten_min_may16[is.na(det_count), det_count := 0]

library(cowplot)
p1 <- ggplot(ten_min_may16, aes(x = interval, y = det_count)) +
  geom_line(color = "forestgreen") +
  geom_vline(xintercept = as.POSIXct("2025-05-16 14:39:00", tz = "America/Chicago"),
             color = "red", linetype = "dashed") +
  labs(title = "Bird Detections — St. Louis Metro (May 16, 2025)",
       x = NULL, y = "Detections per 10 min") +
  theme_minimal()

temp_max <- max(ten_min_may16$mean_temp, na.rm = TRUE)
if (!is.finite(temp_max)) temp_max <- 25

p2 <- ggplot(ten_min_may16, aes(x = interval, y = mean_temp)) +
  geom_line(color = "darkorange") +
  geom_vline(xintercept = as.POSIXct("2025-05-16 14:39:00", tz = "America/Chicago"),
             color = "red", linetype = "dashed") +
  annotate("text", x = as.POSIXct("2025-05-16 14:39:00", tz = "America/Chicago"),
           y     = 30,
           label = "EF3 touchdown\n2:39pm CDT",
           color = "red", hjust = -0.1, size = 3) +
  labs(x = NULL, y = "Temperature (°C)",
       caption = "Data source: BirdWeather via birdweatheR") +
  theme_minimal()

plot_grid(p1, p2, ncol = 1, align = "v")
```

The `get_environment_data()` function retrieves sensor readings for any PUC 
station and time window, enabling researchers to pair environmental context with 
acoustic detections at fine temporal resolution.

---

## Astronomical Events

In addition to environmental sensors, PUCs record light-levels at the sensor, including broadband light levels, near-infrared light, and the ability to isolate bands f1-f8. The combination of acoustic detections and on-board light sensors makes 
BirdWeather data uniquely suited to studying how birds respond to rapid changes 
in light levels — including those caused by solar eclipses.

### The April 8, 2024 Total Solar Eclipse

On April 8, 2024, a total solar eclipse crossed North America along a path 
running from Texas northeast through the Ohio Valley. At stations within the 
path of totality, daylight was completely extinguished for up to four minutes 
around 2:00 pm local time.

The `total_eclipse` dataset contains detections from PUC stations within an 
eclipse path bounding box during April 8, 2024. The `eclipse_light_data` dataset 
contains concurrent light sensor readings from those same stations.

```{r eclipse_load}
data(total_eclipse)
data(eclipse_light_data)
```

```{r eclipse_note, eval = FALSE}
# To reproduce from the API:
# connect_birdweather()
# total_eclipse <- get_detections(
#   from           = "2024-04-08T00:00:00.000Z",
#   to             = "2024-04-09T00:00:00.000Z",
#   ne             = list(lat = 42.639347, lon = -79.672852),
#   sw             = list(lat = 34.331340, lon = -94.570313),
#   confidence_gte = 0.6,
#   station_types  = "puc"
# )
#
# station_ids <- unique(total_eclipse$station_id)
# light_list <- lapply(station_ids, function(sid) {
#   get_light_data(
#     station_id = sid,
#     from       = "2024-04-08T00:00:00.000Z",
#     to         = "2024-04-09T00:00:00.000Z"
#   )
# })
# eclipse_light_data <- data.table::rbindlist(light_list, fill = TRUE)
```

Rather than relying on an external shapefile, we use the PUC light sensor data 
itself to identify stations that experienced totality — defined here as stations 
where broadband light levels dropped below 10% of their morning baseline during 
the eclipse window:

```{r eclipse_totality}
eclipse_light_data[, datetime := as.POSIXct(timestamp,
                    format = "%Y-%m-%dT%H:%M:%S", tz = "America/Chicago")]
total_eclipse[, datetime := as.POSIXct(timestamp,
               format = "%Y-%m-%dT%H:%M:%S", tz = "America/Chicago")]

# Morning baseline (10:00-11:00 am)
station_baselines <- eclipse_light_data[
  datetime >= as.POSIXct("2024-04-08 10:00:00", tz = "America/Chicago") &
  datetime <= as.POSIXct("2024-04-08 11:00:00", tz = "America/Chicago"),
  .(baseline = mean(clear, na.rm = TRUE)), by = station_id]

# Minimum light during totality window (1:45-2:15 pm)
station_min <- eclipse_light_data[
  datetime >= as.POSIXct("2024-04-08 13:45:00", tz = "America/Chicago") &
  datetime <= as.POSIXct("2024-04-08 14:15:00", tz = "America/Chicago"),
  .(min_clear = min(clear, na.rm = TRUE)), by = station_id]

totality_ids <- merge(station_baselines, station_min, by = "station_id")[
  min_clear / baseline < 0.10, station_id]

message(length(totality_ids), " stations identified within path of totality")
```

Of the stations in the bounding box, 25 experienced a light drop of >90%, 
confirming they were likely within the path of totality.

We then compare broadband light levels with Common Grackle (*Quiscalus quiscula*) 
vocalization rates across these stations. Common Grackle is a highly vocal 
diurnal species with peak activity during midday — making it a sensitive 
indicator of unusual light changes:

```{r eclipse_5minbin}
# 5-minute bin helper
bin5 <- function(dt) {
  as.POSIXct(
    paste0(format(dt, "%Y-%m-%d %H:"),
           sprintf("%02d", (as.integer(format(dt, "%M")) %/% 5) * 5),
           ":00"),
    tz = "America/Chicago"
  )
}

# Filter to daytime window and totality stations
eclipse_day_light <- eclipse_light_data[
  station_id %in% totality_ids &
  datetime >= as.POSIXct("2024-04-08 10:00:00", tz = "America/Chicago") &
  datetime <= as.POSIXct("2024-04-08 17:00:00", tz = "America/Chicago")
]

eclipse_day_dets <- total_eclipse[
  station_id %in% totality_ids &
  datetime >= as.POSIXct("2024-04-08 10:00:00", tz = "America/Chicago") &
  datetime <= as.POSIXct("2024-04-08 17:00:00", tz = "America/Chicago") &
  common_name == "Common Grackle"
]

# Bin to 5 minutes
light_bins <- eclipse_day_light[,
  .(mean_clear = mean(clear, na.rm = TRUE)),
  by = .(bin = bin5(datetime))]

det_bins <- eclipse_day_dets[,
  .(det_count = .N),
  by = .(bin = bin5(datetime))]
```

```{r eclipse_plot}
library(cowplot)

p1 <- ggplot(det_bins, aes(x = bin, y = det_count)) +
  geom_line(color = "purple", alpha = 0.2) +
  geom_smooth(color = "purple", se = FALSE, span = 0.15) +
  geom_vline(
    xintercept = as.POSIXct("2024-04-08 13:59:00", tz = "America/Chicago"),
    color = "gray20", linetype = "dashed"
  ) +
  annotate("text",
           x     = as.POSIXct("2024-04-08 13:59:00", tz = "America/Chicago"),
           y     = max(det_bins$det_count, na.rm = TRUE) * 0.95,
           label = "Totality", color = "gray20", hjust = -0.1, size = 3) +
  labs(
    title    = "Common Grackle Vocalizations During the April 8, 2024 Solar Eclipse",
    subtitle = "5-minute detection counts, stations within path of totality",
    x        = NULL,
    y        = "Detections per 5 min"
  ) +
  theme_minimal()

p2 <- ggplot(light_bins, aes(x = bin, y = mean_clear)) +
  geom_line(color = "goldenrod") +
  geom_vline(
    xintercept = as.POSIXct("2024-04-08 13:59:00", tz = "America/Chicago"),
    color = "gray20", linetype = "dashed"
  ) +
  labs(
    subtitle = "Mean broadband clear light across stations in path of totality",
    x        = NULL,
    y        = "Clear Light (counts)",
    caption  = "Data source: BirdWeather via birdweatheR"
  ) +
  theme_minimal()

plot_grid(p1, p2, ncol = 1, align = "v")
```

The light sensor data clearly captures the eclipse — broadband light levels 
collapse to near zero at totality before recovering to normal afternoon levels. 
The concurrent reduction in Common Grackle vocalizations illustrates how 
BirdWeather's integrated acoustic and environmental sensing can be used to 
study behavioral responses to rapid environmental perturbations.

---

## Further Reading

The analyses demonstrated in this vignette represent a small subset of what is 
possible with the BirdWeather dataset. For an example of large-scale ecological 
synthesis using BirdWeather detections, see [Pease & Gilbert (2025)](https://www.science.org/doi/10.1126/science.adv9472), who used 
over 60 million BirdWeather detections across 583 diurnal species to demonstrate 
that light pollution prolongs avian vocal activity by up to 50 minutes in the 
brightest landscapes (*Science*, 387, 848–853).

Additionally, a full analysis of the 2023 Annular and 2024 Total Eclipses using data from the [Eclipse Soundscapes](https://eclipsesoundscapes.org/) project and BirdWeather can be found in [Gilbert et al., 2026](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.73090)

For questions, bug reports, or feature requests, please visit the package 
repository at [github.com/BrentPease1/birdweatheR](https://github.com/BrentPease1/birdweatheR).
